This chapter starts with a discussion regarding the validity of the results and the
possibility of using them to draw more general conclusions. Following this, the results
and what conclusions can be drawn from them are discussed. Finally, suggestions
for future research are given.

\section{Validity of the results}\label{sec:validity}
Two main criticisms concerning the validity of the results can be raised:
\begin{enumerate}
\item Only a few number of queries were used for evaluation;
\item And only one dataset was used for evaluation.
\end{enumerate}

Both of these criticisms will be answered in turn below. Finally a motivation
is provided as to why the results can be considered possible to draw general
conclusions from.

\subsection{Only a few queries were used}
\textit{Only one single query and subsets of that query were used, this is hardly a
  large set of tests for evaluation.}

To answer this criticism it is important to consider the fact that the
possibility for multiple access methods for the same relation depends on the
criteria outlined in Section~\ref{sec:dataset}. The tables fulfilling these
criteria are few, reducing the possibility to use multiple queries, as they will
only involve the same tables anyway.

Furthermore, the query constructed can be considered to be sufficiently complex
to be problematic for the query optimizer, while not being unreasonably
contrived. This means that if the problem does not arise for the query, it is
highly unlikely to do so for a simpler query --- and of little interest if it
does for a more complex query.

So while only one query and subsets of it are used, they cover an interval
of simple to reasonably complex for the query optimizer to handle.

\subsection{Only one dataset was used}\label{sec:validity:dataset}
\textit{Only one dataset was used for evaluation, the results found for the
  problem studied could be very different for another dataset.}

This criticism is one related to the problem of evaluating databases in general:
the performance of the database is often dependent on the dataset used for
testing. While it is correct that the results could vary depending on dataset,
this problem applies to all studies of databases.

Furthermore, the dataset used to test the databases is one taken from the real
world, making it more realistic than those often otherwise used. As such, the
validity of the results of this study are well on par with those of other
studies using less realistic datasets like TPC-H.

So while only one dataset is used, the results found are realistic and well on
par with that of other studies on the subject of databases.

\subsection{Applicability}
The results found in this study do, as all other studies of databases do, suffer
from the problem of the possibility of the results depending on the dataset
used. However, the results found in this study show two things:
\begin{itemize}
\item Different access methods are used to access the same relation;
\item And the behavior of when this is done differs between MariaDB and
  PostgreSQL.\@
\end{itemize}

Neither of these results is tied to the specific dataset used, instead they
both show behavior that can be considered generally applicable to all dataset.

One problem does remain though: because only one dataset is used it is difficult
to answer the extent to which the cardinality estimate affects access method
selection. This is problem is further discussed in the future research section,
Section~\ref{sec:futureresearch}, and suggestions suggested.

\section{Selection of access method}\label{sec:accessmethods}
This section will discuss the main findings of the evaluation, starting with a
discussion regarding the correlation between statistics target and the number of
different access methods used for the same relation. Following this a section
describing the second set of results identified: the correlation between
predicate value and access methods used.

\subsection{The effect of cardinality estimate}
The main purpose of the study was to identify the effect of the cardinality
estimate on the access methods selected. As a way to evaluate this the statistics
target used when analyzing and estimating the cardinality is varied from a low
to a high value. The results found from the evaluation show the following:

\begin{enumerate}
\item PostgreSQL is affected by bad cardinality estimates;
\item PostgreSQL will vary between doing a full table scan or accessing the data
  via an index depending on the cardinality estimate;
\item PostgreSQL becomes more consistent in using the same access method when
  the cardinality is better estimated thanks to an increased statistics target;
\item The cardinality estimate rather than the number of joins seem to be the
  main cause of PostgreSQL's varying access method;
\item MariaDB remains unaffected by the cardinality estimate for the tests done;
\item The primary reason why MariaDB remains unaffected seems to be the fact
  that it will always use an index if one exists, never doing a full table scan.
\item And finally, MariaDB and PostgreSQL will sometimes use different access
  methods, even for a cardinality estimate of high quality.
\end{enumerate}


\subsubsection{PostgreSQL is affected by bad cardinality estimates}
This result is clear from both tests done, shown in
Figure~\ref{fig:plot:eval1:query1} and Figure~\ref{fig:plot:eval1:query2}
respectively, as the number of relations with varying access methods decrease as
the cardinality is better estimated. It is therefore clear from the results that
cardinality estimate will affect PostgreSQL's selection of access method, and do
so to quite some extent --- with four relations having varying access methods
when the statistics target is 1, and decreasing to only one relation at the
default value.

\subsubsection{PostgreSQL only varies between doing a full table scan or using an index}
The output of the tool shows that PostgreSQL only varies between doing a full
table scan or using an index when it varies between several access methods. Even
though it is the case that many of the relations have multiple indexes, as shown
in Table~\ref{table:dataset}, and include non-trivial indexes, as outlined by
the criteria in Section~\ref{sec:dataset}, PostgreSQL never varies between
several indexes. This indicates that even though the cardinality estimate
affects the selection of access method --- it does not do so to such a degree as
to make more than one index considered a viable alternative.

\subsubsection{PostgreSQL is more consistent for better cardinality estimates}
As previously discussed, when the statistics target used for the cardinality
estimate is increased PostgreSQL becomes more consistent in selecting access
method --- resulting in a reduced number of relations with varying access
methods. This behavior can be observed in the result for both queries, shown in
Figure~\ref{fig:plot:eval1:query1} and Figure~\ref{fig:plot:eval1:query2}
respectively.

Furthermore, it is noteworthy that the access method PostgreSQL becomes
consistent in selecting is to use an index rather than a full table scan. This
can be seen for both relations \textit{cmt} and \textit{est}, which vary between
full table scan or an index for a statistics target of $1$. However, when the
target is increased to $d$, both are accessed using only an index.

What can be derived from this is that the correct behavior is to access the
relations using an index, but when the cardinality is badly estimated ---
PostgreSQL will, seemingly incorrectly, consider a full table scan to be better.

\subsubsection{Cardinality estimate is the reason rather than complexity caused
  by joins}
The tests done with the simpler query, query \#2, showed that even though the
query is simpler in terms of the number of joins and tables involved, PostgreSQL
will still vary in what access methods it is using, as can be seen in
Figure~\ref{fig:plot:eval1:query1}. This can be seen as a further motivation that
the cardinality estimate, rather than other factors, is what is causing
PostgreSQL to vary between what access methods it is using.

\subsubsection{MariaDB is unaffected by cardinality estimate}
Unlike PostgreSQL, MariaDB seems to remain unaffected by the cardinality
estimates as can be seen in Figure~\ref{fig:plot:eval1:query1}. It is only one
relation, \textit{ct}, that is accessed with multiple different access methods
and they are the same for both tests, as can seen from the output, shown in.

\subsubsection{MariaDB always use an index if one exists}
Another difference to PostgreSQL is that MariaDB will always use an index if one
exists. This can be seen in the output and more specifically the field
\textit{possible\_keys}, which shows what access methods MariaDB considers
possible candidates. From this it is clear that MariaDB only ever considers
indexes if one exists; for example only one access method is found viable for
\textit{cmm}. This is in clear contrast to PostgreSQL, which varies between two
different methods for \textit{cmm}.

\subsubsection{The access methods vary even with high quality cardinality estimates}
The final discovery that can be drawn from the results is that both databases
use multiple access methods even for high quality estimates of the cardinality.
This can clearly be seen in Figure~\ref{fig:plot:eval1:query1}, where even for a
statistics target of $2d$ there are relations with varying access methods for both
databases.

Furthermore, as MariaDB remains consistent in choosing between the three
different methods regardless of the statistics target used when estimating
cardinality it seems that the choice is a deliberate one. This behavior was
therefore further evaluated by using only $1$ repetition to see if the reason for
varying behavior depended on another factor than varying cardinality estimates.
The results for this evaluation will be discussed in the next section,
Section~\ref{sec:predicatecorrelation}.

\subsection{The effect of predicate value}\label{sec:predicatecorrelation}
The second evaluation done was focused on identifying if a factor other than
cardinality estimate was the reason for the varying use of access methods. This
evaluation was therefore done by using only 1 repetition and so only one
cardinality estimate is the result for all the query plans generated.

The evaluation found the following results:
\begin{enumerate}
\item Both databases will use different access methods for the same estimated
  cardinality;
\item The factor that is the reason for the different access methods used is the
  predicate used to filter rows;
\item And the behavior is not the cause of query complexity.
\end{enumerate}

\subsubsection{Both databases use different access methods}
The first notable result is that both databases will use different access
methods even when the estimated cardinality is the same. This can be
seen in Figure~\ref{fig:plot:eval2:test1} where the relation \textit{ct} for
MariaDB and relations \textit{cmt}, \textit{cmm} and \textit{est} are all
accessed with multiple different access methods.

\subsubsection{The predicate value used is important}
The reason that these relations are accessed in different ways can not be
because of a varying cardinality estimate, as the same estimated cardinality is
used when generating all query plans. The only other thing that varies in the
tests is that all possible values for \sql{:KEY} in the query, seen in
Figure~\ref{fig:sql:query1}, are used to generate query plans. From this it can
be concluded that the the value of \sql{:KEY} in the predicate \sql{ct.key =
  :KEY} will affect the access method for the relations.

It can be further noted that for PostgreSQL the predicate will affect relations
that are not directly involved in it, but also those that are joined together
with it (for example \textit{cmt}). Since MariaDB always use an index if one
exists this behavior is hard to observe as the only relation it will use
different access methods for, \textit{ct}, is the one involved in the predicate.

\subsubsection{The behavior is not the cause of query complexity}
Finally, the behavior of being sensitive to predicate value was evaluated for
two simpler queries to see if the result remains the same when not
joining so many tables together.

The first test was done on query \#2 and the results can be seen in
Figure~\ref{fig:plot:eval2:test2}. The main focus was on evaluating PostgreSQL
to see if it behaved the same and varied access method for the relations
\textit{cmt} and \textit{cmm}. As can be seen in the figure, PostgreSQL
continues to use different access methods.

For MariaDB the behavior was tested using query \#3, which is a simple select
only on the relation \textit{ct}. The results can be seen in
Figure~\ref{fig:plot:eval2:test3} and it is clear that the behavior remains ---
three different access methods are used.

\subsection{Conclusions from the discussion}
This section will discuss what conclusions can be drawn from the discussion
regarding the effect of cardinality and predicate value.

First of all, cardinality estimate has a clear effect on PostgreSQL and the
access methods chosen by its query optimizer. When the statistics target used to
estimate cardinality is low, the number of different access methods used for
relations will vary more than when the statistics target is increased. However,
it will vary between doing a full table scan or using an index, never between
multiple different indexes. This result provides a good indication to the answer
of the question posed at the start of this thesis, showing that cardinality
estimate will affect the access method chosen by PostgreSQL.\@

The effect of cardinality estimate can however only be observed for PostgreSQL,
for MariaDB no such correlation can be seen. The primary reason that MariaDB is
not as sensitive to cardinality estimate is that it never considers doing a full
table scan if it can use an index. However, it is worth noting that it can not
be concluded that MariaDB is unaffected by cardinality estimates. Instead it
might be concluded that it is less likely for MariaDB to vary between different
access methods because of different cardinality estimates, as it must be the
case that it is different indexes it varies between --- rather than an index or
a full table scan, as is the case for PostgreSQL.\@

When evaluating the databases it is also found that another factor that affects
the choice of access method is the predicate value used to filter rows. For
PostgreSQL it affects whether it uses an index or does a full table scan,
whereas it for MariaDB means it will switch between using different indexes.
Since it is usually the case that one index should be the correct one for a
given query this behavior might be seen as unintuitive. Furthermore, this
behavior might cause a query to become slow with the introduction of a seemingly
unrelated index (an example of such behavior is described
in~\cite[Ch.~14]{lahdenmaki_2005_relational_rdidatodossea}).

One final result that is noteworthy is that the two query optimizers behave
considerably different for the tests done. PostgreSQL never considers
many different indexes for the same relation, unlike MariaDB which on the other
hand never considers a full table scan if an index exists, unlike PostgreSQL.\@
This is interesting as it shows that there seems to be no clear best practice
for query optimizers if they differ on such fundamental levels. This also
indicates that further research and evaluation of query optimizer is necessary
in order to find what is the best behavior.

\section{Future research}\label{sec:futureresearch}
This section will cover some suggestions for future research on the topic of
databases, both in general and specifically for the problem of selecting access method.

As discussed in Section~\ref{sec:validity:dataset} one problem when evaluating
databases is the dataset used for evaluation. In this study a dataset based on a
product for the company TriOptima was used to evaluate the databases with a real-world
dataset. However, this dataset can't be made public. Other datasets, like TPC-H
or the more recent JOB, suffer from the problem that they are simpler
than a database used by in the real world; as an example both of them have only one
index per relation, on the primary key.

One important area of research in databases would therefore be to create one or
more realistic datasets featuring complex data, relations and indexes.
These datasets could then be used to provide better and more realistic results
when evaluating databases.

The creation of more datasets would also aid in answering the question posed at
the start of this thesis --- how much the cardinality estimate affects the selection
of access methods. While this thesis can provide an answer for the given dataset
and from this draw some more general conclusions, evaluating a larger number of
datasets would allow more conclusions to be drawn.

The focus of this thesis was the effect of the cardinality estimate on the
access methods used by the query optimizer. The results show that PostgreSQL is
more sensitive to this and will more often do a full table scan when the
cardinality estimate is done with a lower statistics target. MariaDB on the
other hand is observed to be more robust as it never does a full table scan if
an index exists. This leads to the question of which behavior is the best ---
should the query optimizer allow full table scans or is it better to always use
an index if one exists?

This study also show that both database's query optimizers select
different access methods depending on predicate value and that the two do so
differently. This is also unclear if it is correct or if it would be better to
remain consistent in using the same access method.

Finally it can be noted that the results from this study show the need for
further research into query optimizers as two state-of-the-art query optimizers
behave considerably different. The topic of query optimization clearly needs
much further study to arrive at what is best practices.
